---
title: "Weather and Corn Yield Regressions"
author: "Elin Binck"
date: "3/6/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("data/prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")

ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")


```

# Tidy the data

```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)

# convert 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)

```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}

tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)

```

### Winter Temperatures - Winneshiek County

```{r winter temps}

winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)

lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)

```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}

winnewinter$yearsq <- winnewinter$year^2

lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values

ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")

```

### Download NASS corn yield data

```{r yield download}

# set our API key with NASS
nassqs_auth(key = "F470D9A0-22C7-39FB-95F1-5535C03C02B9")

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r}
winneyields <- cornyields %>%
  filter(county_ansi==191) %>%
  group_by(year) %>%
  summarize(meanyield = mean(yield))

ggplot(data = winneyields, aes(x = year, y = meanyield))+
  geom_point()+
  theme_bw()+
  labs(x = "Year", y = "Mean Corn Yield")+
  geom_smooth(method = lm)
```


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r}
winneyields$yearsq <- winneyields$year^2

lm_yieldsquad <- lm(meanyield ~ year + yearsq, winneyields)
summary(lm_yieldsquad)
winneyields$fitted <- lm_yieldsquad$fitted.values

ggplot(winneyields) +
  geom_point(mapping = aes(x = year, y = meanyield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "Year", y = "Mean Corn Yield")
```


### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

There doesn't appear to be a strong relationship between temperature and yields in Winneshiek county.  Adding year improves the model, producing a p-value of 1.46e-111 and and R-squared value of 0.7376. 

```{r}
#join the data together by year
tmaxyields<- winneyields %>% 
  inner_join(winnewinter, by = "year")

#plot the data with a regression line
ggplot(data = tmaxyields)+
  geom_point(aes(x= meantmax, y = meanyield, color = year))+
  geom_smooth(aes(x= meantmax, y = meanyield),method = lm)+
  theme_bw()+
  labs(x = "mean max temp", y = "mean yield")

#create the linear model
lm_tmaxyields <- lm(meanyield ~ meantmax , tmaxyields)

summary(lm_tmaxyields)

#create a model with year as an additional explanatory variable
lm_tmaxyields_yr <- lm(meanyield ~ meantmax + year, tmaxyields)

summary(lm_tmaxyields_yr)
tmaxyields$fitted <- lm_tmaxyields_yr$fitted.values

#create a model with meantmax squared values
tmaxyields$meantmaxsq <- tmaxyields$meantmax^2
lm_tmaxyields_sq <- lm(meanyield ~ meantmax + meantmaxsq, tmaxyields)
summary(lm_tmaxyields_sq)

ggplot(data = tmaxyields)+
  geom_point(aes(x= meantmax, y = meanyield))+
  geom_smooth(aes(x= meantmax, y = fitted), method = lm)+
  theme_bw()+
  labs(x = "mean max temp", y = "mean yield")
  
```


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

There is a significant quadratic relationship between temperature and yield across all counties, so yield was highest when temperatures were in the middle, and lowest when temperatures were either on the low end or the high end.  

```{r}
#filter the corn yeild data to 2018 and convert county_ansi to factor
cornyields2018<-cornyields %>% 
  filter(year == 2018) %>% 
  mutate(across(county_ansi, factor))

#find the meantmax for summer of 2018
all_summer2018 <- tmaxdf %>%
  filter(year==2018 & doy >= 152 & doy <= 243) %>%
  group_by(countyfp) %>%
  summarize(meantmax = mean(tmax)) 

#join the yields with the temp data
all_cornyields2018<-cornyields2018 %>% 
  inner_join(all_summer2018, by = c("county_ansi" = "countyfp"))

#create a linear model with yield and mean tmax
lm_yields2018 <- lm(yield ~ meantmax , all_cornyields2018)
summary(lm_yields2018)

#create a model with yeild, tmax, and tmax squared values
all_cornyields2018$meantmax_sq<-all_cornyields2018$meantmax^2

lm_yields2018_sq <- lm(yield ~ meantmax + meantmax_sq, all_cornyields2018)
summary(lm_yields2018_sq)
all_cornyields2018$fitted <- lm_yields2018_sq$fitted.values

#plot the data
ggplot(data = all_cornyields2018, aes(x = meantmax, y = yield))+
  geom_point()+
  geom_line(aes(x = meantmax, y = fitted))+
  theme_bw()+
  labs(x = "Mean Max Temperature (C)", y = "Corn Yield")
```


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

Including county in the regression reduces the significance of the temperature coefficients, but increases the significance for the relationship as a whole. 

```{r}
#convert county_ansi to factor
cornyields<-cornyields %>% 
  mutate(across(county_ansi, factor))

#find the meantmax for all counties, all years
all_summer <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(countyfp) %>%
  summarize(meantmax = mean(tmax)) 

#join the yields with the temp data
all_cornyields<-cornyields %>% 
  inner_join(all_summer, by = c("county_ansi" = "countyfp"))

#create a tmax squared column
all_cornyields$meantmax_sq<-all_cornyields$meantmax^2

#create a panel regression
lm_all_cornyields_panel<-lm(yield ~ meantmax + meantmax_sq + county_ansi + year, all_cornyields)

summary(lm_all_cornyields_panel)

all_cornyields$fitted<-lm_all_cornyields_panel$fitted.values

ggplot(data = all_cornyields)+
  geom_point(aes(x = yield, y = fitted))+
  geom_smooth(aes(x = yield, y = fitted), method = lm)+
  theme_bw()+
  labs(x = "Yield", y = "Fitted Yield")
```


### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r}
#set parameters for downloading the soy data
params_soy <- list(commodity_desc = "SOYBEANS", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download the data
soyyieldsall <- nassqs_yields(params)

#convert variables to numeric
soyyieldsall$county_ansi <- as.numeric(soyyieldsall$county_ansi)
soyyieldsall$yield <- as.numeric(soyyieldsall$Value)

# clean and filter the data
soyyields <- select(soyyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soyyields <- tibble(soyyields)

#filter to Winneshiek County yields and calculate mean 
winneyields_soy <- soyyields %>%
  filter(county_ansi==191) 

ggplot(data = winneyields_soy, aes(x = year, y = yield))+
  geom_point()+
  theme_bw()+
  labs(title="Winneshiek County", x = "Year", y = "Mean Soy Yield")+
  geom_smooth(method = lm)
```

```


### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.
